A Global Picture of Food Insecurity and Hunger

### 2020-10-13 docs/modeling_fies.tex

Based on various model validation metrics, Random Forest (RF) regression performed the best. The RF regression involves generating a large number of decision trees, each under slightly different conditions and using different sub-samples of the data, and then aggregating the predictions of the decision trees.  Random forest models are known to perform well in the presence of non-linearities and important interactions among predictor variables \citep{breiman2001random}. Since this is likely to be the case in our application, this method is particularly suitable for modeling rates of moderate-to-severe and severe food insecurity.

We observed for the moderate-to-severe as well as the severe RF regression model an in-sample mean squared error (MSE) between 0.002-0.005, with a R-squared (${R}^2$) value of 0.999 (See Fig. \ref{fig:rf_in-sample}). Although RF regression models tackle the issue of overfitting by applying various ensemble methods \citep{breiman2001random}, it is crucial to validate the out-of-sample prediction performance. Hence, we trained the RF regression model with 75\% of the available observations and predicted rates of moderate-to-severe and severe food insecurity for the remaining 25\%. Similarly to the in-sample model validation metrics, the out-of-sample MSE is between 0.008-0.017, with a ${R}^2$ of around 0.98-0.99 for both models (See Fig. \ref{fig:rf_out-sample}).

@@

\subsection{Water Scarcity}
\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/wci_index.png}
  \caption{Water Scarcity Index}
\end{figure}

>>>>

Based on various model validation metrics, Random Forest (RF) regression performed the best. The RF regression involves generating a large number of decision trees, each under slightly different conditions and using different sub-samples of the data, and then aggregating the predictions of the decision trees.  Random forest models are known to perform well in the presence of non-linearities and important interactions among predictor variables \citep{hastie2009elements}. Since this is likely to be the case in our application, this method is particularly suitable for modeling rates of moderate-to-severe and severe food insecurity.

We observed for the moderate-to-severe as well as the severe RF regression model an in-sample mean squared error (MSE) between 0.002-0.005, with a R-squared (${R}^2$) value of 0.999 (See Fig. \ref{fig:rf_in-sample}). To test the out-of-sample prediction performance we used ten-fold cross-validation. First, we partitioned the data into ten equal sized subsamples. Once partitioned, we trained the RF regression models on nine subsamples and tested the models on the remaining one. This process gets repeated ten times, with each subsample being used once for testing. We then calculated the average MSE and ${R}^2$ over all ten out-of-sample predictions. Similarly to the in-sample model validation metrics, the average out-of-sample MSE is between 0.008-0.016, with an average ${R}^2$ of around 0.99 for both models (See Fig. \ref{fig:rf_out-sample}). Besides using the full set of countries to validate the out-of-sample performance, we halved the sample size by excluding high-income countries (HIC) (according to the World Bank definition, REFERENCE NEEDED) and performed ten-fold cross-validation for this limited set of countries. We found that even when excluding HIC the average out-of-sample MSE is between 0.013-0.023, with an average ${R}^2$ of around 0.98 for both models.

@@

\subsection{Water Scarcity Share}
\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/ws_share.png}
  \caption{Water Scarcity Share}
\end{figure}
